---
output:
  html_document: default
  pdf_document: default
---
Loan Data Analysis by Jack Bae
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(lubridate)
library("GGally")
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
loans <- read.csv('/Users/jackbae/Google Drive/Udacity/P4/prosperLoanData.csv', sep = ',', stringsAsFactors=FALSE)
```

## Introduction
***

**Prosper** (www.prosper.com) is a peer-to-peer marketplace lending platform. 
Here is an explanation of Prosper from their website:

> Prosper allows people to invest in each other in a way that is financially and
socially rewarding. On Prosper, borrowers list loan requests between $2,000 and
$35,000 and individual investors invest as little as $25 in each loan listing 
they select. Prosper handles the servicing of the loan on behalf of the matched 
borrowers and investors.

And the process of how peer-to-peer lending at Prosper works:

> - Borrowers choose a loan amount, purpose and post a loan listing.
> - Investors review loan listings and invest in listings that meet their criteria.
> - Once the process is complete, borrowers make fixed monthly payments and 
investors receive a portion of those payments directly to their Prosper account.

The Prosper Loan data analyzed here contains 113,937 records last updated on
03/11/2014.

## Exploring the Dataset

The dataset contains 113937 records with 81 variables.

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(loans)
```

The names of the 81 variables:

```{r echo=FALSE, message=FALSE, warning=FALSE}
names(loans)
```

Overview of dataset variables:

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(loans)
```

While exploring the dataset, I found the need to create new variables and data
frames.
- New variables: ListingCreationYear, ListingCategoryName, CreditScoreAverage
- New data frames: loansFrom2009 (loan data after July 12, 2009),
                   loansTo2009 (loan data before July 12, 2009)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Extract year into new variable
loans$ListingCreationYear = year(loans$ListingCreationDate)
loans$ListingCreationYear <- as.factor(loans$ListingCreationYear)

# Match numeric loan type to correct labels
alphaLoanType <- c("Not Applicable", "Debt Consolidation", "Home Improvement",
           "Business", "Personal Loan", "Student Use", "Auto", "Other",
           "Baby&Adoption", "Boat", "Cosmetic Procedure", "Engagement Ring", 
           "Green Loans", "Household Expenses", "Large Purchases",
           "Medical/Dental", "Motorcycle", "RV", "Taxes", "Vacation", 
           "Wedding Loans")
loans$ListingCategoryName <- factor(loans$ListingCategory, 
                                    levels = seq(0:20), labels = alphaLoanType)

# Create a new variable of getting a single value for credit score by 
# getting average of lower and upper limits of credit score range.
loans$CreditScoreAverage = ceiling((loans$CreditScoreRangeLower + 
                              loans$CreditScoreRangeUpper) / 2)

# Create new data frames: loan data before July 12, 2009 and loan data after
# July 12, 2009.
loansFrom2009 <- subset(loans, loans$ListingCreationDate > "2009-07-12")
loansTo2009 <- subset(loans, loans$ListingCreationDate <= "2009-07-12")


```

# Univariate Plots Section

### Is there a trend in the number of Prosper loans?

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = ListingCreationYear), data = loans) + 
  geom_bar() + 
  ggtitle("Number of Loans by Year") +
  xlab("Loan Creation Year")
```

The number of loans by year graph shows that the number of loans increased from
2006 to 2009 but decreased in 2009 due to the global financial crisis. Loans
increased again starting in 2010 through 2013. The decrease in loans can be 
possibily explained by not having full year data for 2014.

```{r echo=FALSE, message=FALSE, warning=FALSE}
range(loans$ListingCreationDate)
```

Checking the range of loan listing creation date, the data confirms that the
most recent data for 2014 is March 10, confirming 2014 data does not contain 
full year data.

### What is the distribution of loans by credit grade?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Reorder Credit Grade in descending order
loansTo2009$CreditGrade <- ordered(loansTo2009$CreditGrade, levels = 
                                     c("AA","A","B","C","D",
                                       "E","HR","NC"))

ggplot(aes(x = CreditGrade), data = loansTo2009) + 
  geom_bar() + 
  ggtitle("Number of Loans by Credit Grade (Pre-2009)") +
  xlab("Credit Grade")
```

I was interested in investigating who uses peer-to-peer lending platform. One
way of looking the data was throuth the credit grade that Prosper created. 
The credit grade data for data before July 12, 2009 shows a normal distribution
of data. Prosper moved to a new rating system starting July 12, 2009.

```{r echo=FALSE, message=FALSE, warning=FALSE}
loansFrom2009$ProsperRating..Alpha. <- ordered(loansFrom2009$ProsperRating..Alpha.,
                                               levels = c("AA","A","B","C","D",
                                       "E","HR","NC","NA"))
ggplot(aes(x = ProsperRating..Alpha.), data = loansFrom2009) + 
  geom_bar() + 
  ggtitle("Number of Loans by Credit Grade (Post-2009)") +
  xlab("Prosper Rating")
```

Post-2009 credit grade has been renamed as Prosper Rating. Although it is showing
a normal distribution similar to pre-2009 credit grade distribution, the Prosper
Rating is showing fewer applicants with the highest and the lowest distribution.

### Distribution of Credit Scores of Prosper Applicants

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = CreditScoreAverage), data = loans) + 
  geom_bar() + 
  ggtitle("Average Credit Score") +
  xlab("Credit Score Average")

summary(loans$CreditScoreAverage)
```

In further understanding Prosper applicants, I have plotted credit score of 
applicants. The dataset provided fields of lower and upper credit score range.
The data here is the average of the lower and upper range. 

According to study by [ValuePenguin](https://www.valuepenguin.com/average-credit-score),
the average American credit score is 695, which is almost identical the score of
695.6 of Prosper applicants. This shows that Prosper applicants are not skewed
towards either low or high credit score applicants. THe normal distribution of 
the plot further supports this finding.

### Distribution of risk score built by Prosper

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = ProsperScore), data = loansFrom2009) + 
  geom_bar() + 
  scale_x_continuous(breaks=seq(0, 12, 1)) +
  ggtitle("Prosper Score (Post-2009)")  +
  xlab("Prosper Score")

summary(loans$ProsperScore)
```

The Prosper Score is a custom risk score built using historical Prosper data.
The higher Prosper Score indicates the lowest risk users. The Prosper Score 
shows a different distribution from Prosper Rating data. This indicates the
Prosper Score is calculated based on multiple factors.

### Distribution of loan term

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = as.factor(Term)), data = loans) + 
  geom_bar() + 
  ggtitle("Loan Term") +
  xlab("Term")
table(loans$Term)
```

After learning who used Prosper's service, I was interested in how the Prosper
lending service was used. The distribution of loan term shows that approximately
77% of the loans were 36 month loans. It is followed by 21.5% of 60 month loans 
and 1.5% of loans being 12 month loans.

### Distribution of Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE}
loans$LoanStatus <- ordered(loans$LoanStatus,levels = c("Cancelled","Chargedoff",
                                                        "Completed","Current",
                                                        "Defaulted",
                                                        "FinalPaymentInProgress",
                                                        "Past Due (1-15 days)",
                                                        "Past Due (16-30 days)",
                                                        "Past Due (31-60 days)",
                                                        "Past Due (61-90 days)",
                                                        "Past Due (91-120 days)",
                                                        "Past Due (>120 days)"))

ggplot(aes(x = reorder(LoanStatus, LoanStatus, function(x) -length(x))), 
           data = loans) + 
  geom_bar() + 
  ggtitle("Loan Status") +
  scale_y_sqrt() +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  xlab("Loan Status")
```

Next, I was interested in the current status of Prosper loans. Most of the loan 
status is current. It is followed by completed, charged off, and defaulted loan
status.

### Distribution of Loans by Listing Category

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = ListingCategoryName), data = loans) + 
  geom_bar() + 
  ggtitle("Number of Loans by Listing Category") +
   scale_y_log10() +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  xlab("Listing Category")
```

Observing the plot, the most common category the borrower selected when posting
their listing is auto. It is followed by debt consolidation and home improvement.

### Number of Current Credit Lines

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = CurrentCreditLines), data = loans) + 
  geom_bar() + 
  ggtitle("Number of Current Credit Lines") +
  xlab("Current Credit Lines")

summary(loans$CurrentCreditLines)
```

The number of current credit lines at the time when credit profile was pulled
ranges from 0 to 59, with average of 10.32.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = OpenCreditLines), data = loans) + 
  geom_bar() + 
  ggtitle("Number of Loans by Open Credit Lines") +
  xlab("Open Credit Lines")
summary(loans$OpenCreditLines)
```

The number of open credit lines at the time when credit profile was pulled shows
a similar distribution with current credit lines. The number of loans by open
credit lines ranges from 0 to 54, with average of 9.26.

### Number of Loans by Income Range

```{r echo=FALSE, message=FALSE, warning=FALSE}
loans$IncomeRange <- ordered(loans$IncomeRange,levels = c("Not displayed",
                                                          "Not employed",
                                                          "$0","$1-24,999",
                                                          "$25,000-49,999",
                                                          "$50,000-74,999",
                                                          "$75,000-99,999",
                                                          "$100,000+"))

ggplot(aes(x = IncomeRange), data = loans) + 
  geom_bar() + 
  ggtitle("Number of Loans by Income Range") +
  theme(axis.text.x=element_text(angle=90,hjust=1)) 
```

The income range data for borrowers show that the $25,000-49,999 income bracket
request most loans, followed by $50,000-74,999 income range.

### Distribution of Total Prosper Loans

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = TotalProsperLoans), data = loans) + 
  scale_x_continuous(breaks=seq(0, 10, 1)) +
  geom_bar() + 
  ggtitle("Total Prosper Loans") 
```

Data shows that most of the Prosper applicants were first-time borrowers. This
was to be expected because of Prosper's short history (founded in 2005).

### Distribution of Loan by Amount

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginalAmount), data = loans) + 
  geom_histogram(binwidth=500) + 
  ggtitle("Number of Loans by Original Loan Amount") 
summary(loans$LoanOriginalAmount)
```

The distribution of loan amount shows a median amount of $6,500. The plot shows
significant number of loans were increments of $5,000 with plots showing spikes
in loan amount of $5,000, $10,000, $15,000, and $20,000.

### Number of Loans by Originial Loan Amount

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = MonthlyLoanPayment), data = loans) + 
  geom_histogram(binwidth = 20) +
  ggtitle("Number of Loans by Monthly Loan Payment") 
summary(loans$MonthlyLoanPayment)
```

The distribution of monthly loan payment shows median of $217.7. The plot shows
most loan payment under $500 per month. With maximum payment of $2,251, this 
plot also shows flexibility in loan payment, which could be one factor in 
using Prosper loan service.


# Univariate Analysis

### What is the structure of your dataset?

The Prosper loan dataset contains 113,937 loan data across 81 variables. 

### What is/are the main feature(s) of interest in your dataset?

In investigating the Prosper loan dataset, I was interested in who uses 
peer-to-peer lending service. The univariate plots revealed that many high income
people used Prosper loan service. I was surprised and intrigued by this trend. 
I would like to investigate further categorize which people are more likely to 
take more loans and for what reason.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

In analysis of the dataset, I found the need to create a listing name for 
ListingCategory, where it originally had numerical values. Also, I found that
the ListingCatoryYear would be helpful in seeing trends in listings year to year
and CreditScoreAverage variable to have a single reference point for credit score
instead of having a lower and upper range.

### Did you create any new variables from existing variables in the dataset?

I created ListingCreationYear, ListingCategoryName, and CreditScoreAverage
variables to the dataset in assisting analyzing Prosper loan dataset.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

It was interesting to see how much the 2009 Global financial crisis had impacted
the number of Prosper loans initiated that year. I created ListingCreationYear to
show prosper loan by year instead of showing continuous data variable.

# Bivariate Plots Section

### Prosper Rating vs Credit Score Average

```{r echo=FALSE, message=FALSE, warning=FALSE}
loansFrom2009$ProsperRating..Alpha. <- ordered(loansFrom2009$ProsperRating..Alpha.,
                                               levels = c("AA","A","B","C","D",
                                       "E","HR","NC","NA"))

ggplot(aes(x=ProsperRating..Alpha., y=CreditScoreAverage), 
       data=loansFrom2009) +
    geom_boxplot() +
  ggtitle("Prosper Rating vs Credit Score Average") +
  xlab("Prosper Rating") +
  ylab("Credit Score Average")

by(loans$CreditScoreAverage, loans$ProsperRating..Alpha., summary)
```

As expected, the higher Prosper Rating is associated with a higher average credit
score. However, it was surprising to learn that even the lowest Prosper Ratings
of D and E had applicants with average credit score of 870. This fact shows that
Prosper Rating is based on multiple factors.

### Prosper Rating vs Loan Original Amount

```{r echo=FALSE, message=FALSE, warning=FALSE}
loansFrom2009$ProsperRating..Alpha. <- ordered(loansFrom2009$ProsperRating..Alpha.,
                                               levels = c("AA","A","B","C","D",
                                       "E","HR","NC","NA"))

ggplot(aes(x=ProsperRating..Alpha., y=LoanOriginalAmount), 
       data=loansFrom2009) +
    geom_boxplot() +
  ggtitle("Prosper Rating vs Loan Original Amount") +
  xlab("Prosper Rating") +
  ylab("Loan Original Amount")

by(loans$LoanOriginalAmount, loans$ProsperRating..Alpha., summary)
```

The plot shows correlation between Prosper Rating and loan amount of applicants.
Data shows that applicants with higher rating tend to have higher loans on 
Prosper.

### Listing Category vs Loan Original Amount

```{r echo=FALSE, message=FALSE, warning=FALSE}
loansFrom2009$ProsperRating..Alpha. <- ordered(loansFrom2009$ProsperRating..Alpha.,
                                               levels = c("AA","A","B","C","D",
                                       "E","HR","NC","NA"))

ggplot(aes(x=ListingCategoryName, y=LoanOriginalAmount), 
       data=loansFrom2009) +
    geom_boxplot() +
  ggtitle("Listing Category vs Loan Original Amount") +
  xlab("Listing Category") +
  ylab("Loan Original Amount") +
  theme(axis.text.x=element_text(angle=90,hjust=1)) 
```

The Prosper Rating vs. Loan Original Amount plot shows home improvement category
with the highest average loan amount and personal loan category with the lowest
average loan amount. 

### Home Owner vs Loan Original Amount

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=IsBorrowerHomeowner, y=LoanOriginalAmount), 
       data=loans) +
    geom_boxplot() +
  ggtitle("Home Owner vs Loan Original Amount") +
  xlab("Home Owner") +
  ylab("Loan Original Amount") +
  theme(axis.text.x=element_text(angle=90,hjust=1)) 
```

The data shows that homeowners have, on average, higher loan amount from Prosper.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

In investigating listing category vs. loan original amount data, I wass surprised
to see listing category did not have a significant correlation with loan original
amount. I expected greater variance with some category than others.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I observed that homeowner status has strong correlation with loan original amount.

### What was the strongest relationship you found?

The strongest relationship was prosper rating vs. credit score average. It was
still surprising to learn that some people with high credit score still rated
low on Prosper rating.

# Multivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE}
creditsubset <- c("CreditScoreAverage","CurrentCreditLines","OpenCreditLines",
                  "TotalCreditLinespast7years","OpenRevolvingAccounts",
                  "InquiriesLast6Months","TotalInquiries","CurrentDelinquencies",
                  "DebtToIncomeRatio","IncomeRange","ProsperRating..numeric.")
loanscredit <- loans[creditsubset]

ggcorr(loanscredit, label = TRUE, label_size = 3,
       hjust = 0.8, size = 2.5, color = "black", layout.exp = 2)
```

I was interested in learning which factors in the dataset has greatest correlation
with each other and ultimately with the Prosper Rating. The heat map shows that 
the Pearson correlation between chosen variables for investigation. The Prosper
Rating has greatest correlation with average credit score. It has little correlation
with number of current credit lines, number of open credit lines, and number of
revolving accounts. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=BorrowerAPR, y = LoanOriginalAmount), data = loans) +
  ggtitle("Borrower APR vs Loan Original Amount by Prosper Rating")  +
  geom_point(color="black",size=0.1) +
  facet_wrap(~ProsperRating..Alpha., ncol=2,  scales = "fixed")

```

In this plot, I wanted to investigate how much loans people take depending on
Prosper rating. As expected, higher credit raters enjoy lower APR. The plot
shows that higher credit raters tends to take higher loans, while lower credit
raters has more variance in APR even within Prosper rating bracket.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In investigating subset of Prosper loan data, I was able to find that the average
credit score had the strongest correlation.

In investigating Borrower APR vs Loan Original Amount by Prosper Rating I found
that people with higher Prosper Rating are likely to have larger loans. I also 
found that the lower Prosper Rating users see more variation in APR even within
their Prosper Rating bracket.

### Were there any interesting or surprising interactions between features?

I was surprised to see little correlation betewen number of credit accounts
opened and Prosper Rating.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = ListingCreationYear), data = loans) + 
  geom_bar() + 
  ggtitle("Number of Loans by Year") +
  xlab("Loan Creation Year")
```

### Description One

This plot shows not only shows the growth of Prosper's business but also the 
growth of peer-to-peer lending service. It is showing steady growth since 2005,
with the exception of Global Finanical Crisis in 2009.

### Plot Two
```{r echo=FALSE, Plot_Two}
loansFrom2009$ProsperRating..Alpha. <- ordered(loansFrom2009$ProsperRating..Alpha.,
                                               levels = c("AA","A","B","C","D",
                                       "E","HR","NC","NA"))

ggplot(aes(x=ProsperRating..Alpha., y=CreditScoreAverage), 
       data=loansFrom2009) +
    geom_boxplot() +
  ggtitle("Prosper Rating vs Credit Score Average") +
  xlab("Prosper Rating") +
  ylab("Credit Score Average")
```

### Description Two

I was interested to see how Prosper rating is calculated. Although the Prosper
Rating is strongly correlated with credit scores, I was also surprised to see
lower Prosper Rating associated with high credit scores. This shows that Prosper
Rating is calculated based on multiple factors.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggcorr(loanscredit, label = TRUE, label_size = 3,
       hjust = 0.8, size = 2.5, color = "black", layout.exp = 2)
```

### Description Three

In continuation of investigating how Prosper Rating is calculated, I created a
subset of data with the factors that could potentially have correlation with
Prosper Rating. The credit score has the strongest correlation to Prosper Rating.

------

# Reflection

Investigating the Prosper Loan dataset, I gained insight into the peer-to-peer
lending service. I had no idea on who used this service and the APR the users
enjoyed. In investigating the dataset, I was most interested in who uses
Prosper's service, how they used the service, and for what purpose. The biggest
struggle was first with the ggplot syntax and organizing plots so that it was
easy for the user to interpret the data. The second major challenge for me was
finding interesting trends from the dataset. Finding trends and relationships 
from 81 variables was a challenge with the number of variables and data.
I found that by dividing the variables into similar and smaller pieces, I was 
able to decipher data more easily.

One interesting future work could be finding trends that could influence Prosper
Rating. It is likely that the current Prosper Rating does not account for 
applicants based on economy situation. During a recession like 2009, there could 
be factors more relevant to risk factors than during an economic boom. Having a
dynamic Prosper Rating based on economic sitaution could help filter applicants
effectively and ultimately bring more profits to Prosper.